Electricity transmission expansion has suffered many delays in Europe in recent
decades, despite its significance for integrating renewable electricity into the
energy system. A hydrogen network which reuses the existing fossil gas network
would not only help supply demand for low-emission fuels, but could also help to
balance variations in wind and solar energy across the continent and thus avoid
power grid expansion. We pursue this idea by varying the allowed expansion of
electricity and hydrogen grids in net-zero \co scenarios for a sector-coupled
European energy system with high shares of renewables and self-sufficient
supply. We cover the electricity, buildings, transport, agriculture, and
industry sectors across 181 regions and model every third hour of a year. With
this high spatio-temporal resolution, the model can capture bottlenecks in
transmission networks, the variability of demand and renewable supply, as well
as regional opportunities for the retrofitting of legacy gas infastructure and
geological hydrogen storage. Our results show a consistent system cost
reductions with a pan-continental hydrogen network that connects regions with
low-cost and abundant renewable potentials to demand centers, synthetic fuel
production and cavern storage sites. Developing a hydrogen network reduces
system costs by up to \maxhybenefitrel\%, with highest benefits when electricity
grid reinforcements cannot be realised. Between \minretroshare\% and
\maxretroshare\% of this network could be built from repurposed natural gas
pipelines. However, we find that hydrogen networks can only partially substitute
for power grid expansion, and that while both can achieve strongest cost savings
of \gridbenefitrel\% together, neither is truly essential if such a cost premium can
be accepted to achieve alternate goals.

% TODO 150 words max

% Our findings show that many cost-effective infrastructure options exist for
% energy transport in Europe, allowing
% policymakers to choose compromise energy system designs.